---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data
Let us first load the data into a data frame.

```{r,echo = TRUE}
df <- read.csv(file = "activity.csv")
```

## What is mean total number of steps taken per day?
Now let us obtain step totals on a daily basis in a data frame. To do that, first load in the sqldf package.

```{r,echo = TRUE}
library(sqldf)
daytotals <- sqldf("SELECT date,SUM(steps) as total_steps from df WHERE steps IS NOT(NULL) GROUP BY date")
print(daytotals)
```

For a histogram of daily total steps, we have the following R code.

```{r,echo = TRUE}
hist(daytotals$total_steps,main = "Histogram of Total Steps",xlab = "Daily Total Steps")
```

From the code chunk below, we can obtain that mean and median of the total number of daily steps taken.

```{r,echo = TRUE}
print(mean(daytotals$total_steps))
print(median(daytotals$total_steps))
```

## What is the average daily activity pattern?
Now for the average daily activity pattern, we have that the following pattern.

```{r,echo = TRUE}
intervalmeans <- sqldf("SELECT interval,AVG(steps) as mean_steps from df WHERE steps IS NOT(NULL) GROUP BY interval")
plot(y = intervalmeans$mean_steps,x = intervalmeans$interval,type = "l",xlab = "interval",ylab= "mean steps",main = "Mean ")
```

From the code chunk below, we get that the 167th 5-minute interval i.e interval marked as 835 has the highest average daily mean steps.

```{r,echo = TRUE}
max_mean_steps <- max(intervalmeans$mean_steps)
subset(x = intervalmeans,intervalmeans$mean_steps == max_mean_steps,select = 1)
```

## Imputing missing values
First subset the data that has missing values and then count the number of missing values in that datframe.

```{r,echo = TRUE}
missingdf <- subset(df,is.na(df$steps))
length(missingdf$steps) # Hence 2304 missing values
```

Consider replacing missing values with the mean of values for that time interval over all days.

```{r,echo = TRUE}
imputed_df <- df
length_df <- nrow(imputed_df) # Length of the data frame
for(i in 1:length_df){
  if(is.na(imputed_df[i,1])){
    imputed_df[i,1] <- subset(intervalmeans,intervalmeans$interval == imputed_df[i,3])[1,2]
  }
}
```

Following is the histogram of the imputed data frame.

```{r,echo = TRUE}
imputedtotals <- sqldf("SELECT date,SUM(steps) as total_steps from imputed_df GROUP BY date") # This data frame has daily step totals
hist(imputedtotals$total_steps,main = "Histogram of Imputed Total Steps",xlab = "Daily Imputed Total Steps")
print(mean(imputedtotals$total_steps))
print(median(imputedtotals$total_steps))
```

It can be seen from above that mean of total daily steps hasn't changed even with imputation which can be easily proven.
However, the median has slightly moved higher as a result of imputation.

## Are there differences in activity patterns between weekdays and weekends?
Use the imputed data frame to assign a factor variable for weekdays.

```{r,echo = TRUE}
library(timeDate) # This package has useful functions for time-date conversions
imputed_df$category <- isWeekday(as.Date(imputed_df$date))
for(i in 1:length_df){
  if(imputed_df[i,4]){
    imputed_df[i,4] <- "weekday"
  }
  else{
    imputed_df[i,4] <- "weekend"
  }
}
```

The following code creates the required time series panel plot. 

```{r,echo = TRUE}
library(ggplot2)
imputed_intervalmeans <- sqldf("SELECT interval,category, AVG(steps) as mean_steps from imputed_df WHERE steps IS NOT(NULL) GROUP BY category,interval")
gplt <- ggplot(imputed_intervalmeans, aes(x = interval,y = mean_steps))+geom_line()+facet_grid(category ~.)
print(gplt)
```

Hence it can be seen that there are differences in activity patterns between weekdays and weekends.